---
title: "Class 7: Machine Learning 1"
author: "Nathan Phan (PID: A17395036)"
format: gfm
---

Today we will explore some fundamental machine learning methods including clustering and dimensionality reduction.

## K-means clustering

To see how this works let's first makeup some data to cluster where we kow what the answer should be

```{r}
hist( rnorm(500, mean=5) )
```

```{r}
x <- c(rnorm(30, mean=-3), rnorm(30, mean=3) )
y <- rev(x)
```


*cbind binds by columns*
```{r}
x <- cbind(x,y)
plot(x)
```

The function for K-means clustering in "base" R is `kmeans()`

```{r}
k <- kmeans(x, centers = 2)
k
```

To get at the resukts of the returned list object we can use the dollar `$` syntax

> Q. How many points are in each cluster

```{r}
k$size
```

> Q. What 'component' of your result object details
  - cluster assignment/membership?
  - cluster center?
  
```{r}
k$cluster
k$centers
```

> Q. Make a clustering results sfigure of the data colored by cluster membership and show cluster centers

```{r}
plot(x, col=c("red", "blue"))
```

```{r}
plot(x, col=k$cluster, pch=16)
points(k$centers, col="orange", pch=15, cex=2)
```

```{r}
k4 <- kmeans(x, centers=4)
plot(x, col=k4$cluster)
points(k4$centers, pch=15)
```


Screen plot to pick k `centers` value

brute-force
```{r}
k1 <- kmeans (x, centers=1)
k2 <- kmeans (x, centers=2)
k3 <- kmeans (x, centers=3)
k4 <- kmeans (x, centers=4)
k5 <- kmeans (x, centers=5)
```


```{r}
z <- c(k1$tot.witness,
  k2$tot.withinss,
k3$tot.withinss,
k4$tot.withinss,
k5$tot.withinss)
plot(z, typ="b")
```

```{r}
n <- NULL
for(i in 1:5 ) {
  n <- c(n, kmeans(x, centers=i)$tot.withinss)
}

plot(n, typ="b" )
```

## Hierarchical Clustering

The main "base" R function for Hierarchical Clustering is called `hclust()`. Here we can't just input our data we need to first calculate a distance matrix (e.g. `dist()`) for our data and use this as input to `hclust()`

```{r}
d <- dist(x)
hc <- hclust(d)
hc
```

There is a plot method for hclust results lets try it

```{r}
plot(hc)
abline(h=8, col="red")
```

To get our cluster "membership" vector (i.e. our main clustering result) we can "cut" the tree at a given height or at a height that yields a given "k" groups.

```{r}
cutree(hc, h=8)

```

```{r}
grps <- cutree(hc, k=2)
```

> Q. Plot the data with our hclust result coloring

```{r}
plot(x, col=grps)
```

##Principal Component Analysis (PCA)

## PCA of UK food data

import food data from an online CSV file:

```{r}
url <- "https://tinyurl.com/UK-foods"
x <- read.csv(url)
head(x)
```


```{r}
x <- read.csv(url, row.names = 1)
x
```

Some base figures

```{r}
barplot(as.matrix(x), beside=T, col=rainbow(nrow(x)))
```

There is one plot that can be useful for small datasets:

```{r}
pairs(x, col=rainbow(nrow(x)), pch=16)
```

> Main point: It can be difficult to spot major trends and patterns even in retaively small multivariate datasets (here we on have 17 dimensions, typically we have 1000s).

## PCA to the rescue

The main function in "base" R for PCA is called `prcomp()`


I will take the transpose of our food data so the "foods" are in the columns:


```{r}
pca <- prcomp(t(x))
summary(pca)
```

```{r}
pca$x
cols <- c("orange", "blue", "red", "green")
plot(pca$x[,1], pca$x[,2], col=cols, pch=16)
```

```{r}
library(ggplot2)
```


```{r}
ggplot(pca$x) +
  aes(PC1, PC2) +
  geom_point(col=cols)
```


```{r}
ggplot(pca$rotation) +
  aes(PC1, rownames(pca$rotation)) +
  geom_col()
```

PCA looks super useful and we will come back to describe this further next day :-)














