---
title: "Class10"
author: "Nathan Phan (PID: A17395036)"
format: gfm
toc: True
---

## Backgound
As it is nearly halloween and the half way point in the quarter let's do a mini project to help us figure out the best candy!

Our come from the 538 website and is available as a CSV file:

## Data Import

```{r}
candy <- read.csv("candy-data.csv", row.names = 1)
head(candy)
```


>Q1. How many different candy types are in this dataset?

```{r}
nrow(candy)
```
There are 85 different types of candies.

```{r}
candy|>
  nrow()
```

```{r}
library(tidyverse)
candy%>%
  nrow()
```

>Q2. How many candy types are in the dataset?

```{r}
sum(candy$fruity==1)
```
There are 38 types of fruity candy in teh dataset

>Q3. What is your favorite candy in the dataset and what is it's winpercent value?

My favorite candy is Twix!

```{r}
candy["Twix", ]$winpercent
```

```{r}
library(dplyr)

candy|>
  filter(rownames(candy)=="Twix") |>
  select(winpercent)
```
>Q4. What is the winpercent value for “Kit Kat”?

```{r}
candy["Kit Kat", ]$winpercent
```
The winpercent value is 76.7686

>Q5. What is the winpercent value for “Tootsie Roll Snack Bars”?

```{r}
candy["Tootsie Roll Snack Bars", ]$winpercent
```
The winpercet is 49.6535

## Quick overview of the dataset

```{r}
skimr::skim(candy)

```

>Q6.Is there any variable/column that looks to be on a different scale to the majority of the other columns in the dataset?

The winpercent is on a 0-100 scale the rest are 0-1 scale

>Q7. What do you think a zero and one represent for the candy$chocolate column?

That the candy does not contain chocolate

>Q8. Plot a histogram of winpercent values

```{r}
library(ggplot2)

ggplot(candy, aes(x = winpercent)) +
geom_histogram(binwidth = 5, fill = "orange", color = "black") +
labs(title = "Distribution of Candy Win Percentages",
x = "Win Percent",
y = "Count")

```


>Q9. Is the distribution of winpercent values symmetrical?

```{r}
ggplot(candy)+
  aes(winpercent) +
  geom_density()

```
No the distribuition is not symmetrical.

>Q10. Is the center of the distribution above or below 50%?

```{r}
mean(candy$winpercent)
```
```{r}
summary(candy$winpercent)
```


>Q11. On average is chocolate candy higher or lower ranked than fruit candy?

```{r}
# 1. Find all chocolate candy in the dataset
# 2. Find their winpercent values
# 3. Calculate the mean of these values

# 4-6. Do the same for fruity candy
# 7. Compare mean winpercents of chocolate vs fruity
# 8. Pick the highest as the winner

mean(candy[candy$chocolate==1,]$winpercent)
choc.ind <- candy$chocolate==1
choc.win <- candy[choc.ind,]$winpercent
choc.mean <- mean(choc.win)
```

```{r}
mean(candy[candy$chocolate==1,]$winpercent)
```


```{r}
mean(candy[candy$fruity==1,]$winpercent)
fruit.ind <- candy$fruity==1
fruit.win <- candy[fruit.ind,]$winpercent
fruit.mean <- mean(fruit.win)
``` 

>Q12. Is this difference statistically significant?

```{r}
t.test(choc.win, fruit.win)
```

## Overall Candy Rankings

>Q13. What are the five least liked candy types in this set?

```{r}
candy |>
  arrange(winpercent) |>
  head(5)
```

```{r}
ord.ind <- order(candy$winpercent)
head(candy[ord.ind,],5)
```

>Q14. What are the top 5 all time favorite candy types out of this set

```{r}
candy|>
  arrange(-winpercent) |>
  head(5)
```

>Q15. Make a first barplot of candy ranking based on winpercent values.

```{r}
ggplot(candy) +
  aes(winpercent, rownames(candy)) +
  geom_col()
```

>Q16. This is quite ugly, use the reorder() function to get the bars sorted by winpercent?

```{r}
ggplot(candy) +
  aes(x=winpercent,
      y=reorder(rownames(candy), winpercent)) +
  geom_col()
```

Add some color based on the "type of candy"

```{r}
my_cols <- rep("black", nrow(candy))
my_cols[as.logical(candy$chocolate)] <- "chocolate"
my_cols[as.logical(candy$fruity)] <- "pink"
my_cols[as.logical(candy$bar)] <- "brown"
my_cols
```

```{r}
ggplot(candy) +
  aes(x=winpercent,
      y=reorder(rownames(candy), winpercent)) +
  geom_col(fill=my_cols)
```

## Winpercent and Pricepercent

A plot with both variables/columns winpercent and pricepercent

```{r}
my_cols[as.logical(candy$fruity)] <- "red"

ggplot(candy) +
  aes(x= winpercent,
      y= pricepercent,
      label=rownames(candy)) +
  geom_point(col=my_cols) +
  geom_text()


```

```{r}
library(ggrepel)

ggplot(candy) +
  aes(x= winpercent,
      y= pricepercent,
      label=rownames(candy)) +
  geom_point(col=my_cols) +
  geom_text_repel(max.overlaps = 7)



```

## Exploring the correlation structure

Now that we’ve explored the dataset a little, we’ll see how the variables interact with one another. We’ll use correlation and view the results with the corrplot package to plot a correlation matrix.

```{r}
library(corrplot)

cij <- cor(candy)
corrplot(cij)
```

## Principal Component Analysis

The function to use is called `prcomp()` with an optinoal `scale=T/F` argument.

```{r}
pca <- prcomp(candy, scale=TRUE)
summary(pca)
```

Our main PCA result figure

```{r}
ggplot(pca$x) +
  aes(PC1, PC2, label=rownames(pca$x)) +
  geom_point(col=my_cols)+
  geom_text_repel(col=my_cols)
```

We should also examine the varaible "loadings" or contribuitions of the origional variables to the new PCs

```{r}
pca$rotation
```

```{r}
ggplot(pca$rotation) +
  aes(PC1, rownames(pca$rotation)) +
  geom_col()
```


```{r}
p <- ggplot(pca$x) +
  aes(PC1, PC2, label=rownames(pca$x)) +
  geom_point(col=my_cols)+
  geom_text_repel(col=my_cols)
```

Interactive plots that can be zoomed on and "brushed" out

```{r}
library(plotly)

#plotly(p)
```

>Q17: worst ranked chocolate candy (lowest winpercent among chocolate==TRUE)

```{r}
choc_df <- subset(candy, chocolate == TRUE)
worst_choc <- choc_df[order(choc_df$winpercent), , drop=FALSE]
head(worst_choc, 1)

```

>Q18. What is the best ranked fruity candy?

```{r}
fruity_df <- subset(candy, fruity == TRUE)
best_fruit <- fruity_df[order(-fruity_df$winpercent), , drop=FALSE]
head(best_fruit, 1)
```


>Q19. Which candy type is the highest ranked in terms of winpercent for the least money - i.e. offers the most bang for your buck?

```{r}
candy$bang_ratio <- candy$winpercent / (candy$pricepercent + 1e-6)
candy[order(-candy$bang_ratio), ][1:5, c("winpercent","pricepercent","bang_ratio")]

```

>Q20. What are the top 5 most expensive candy types in the dataset and of these which is the least popular?

```{r}
ord <- order(candy$pricepercent, decreasing = TRUE)
top5_expensive <- candy[ord, c("pricepercent","winpercent")][1:5, ]
top5_expensive
top5_expensive[which.min(top5_expensive$winpercent), ]

```

>Q22. Examining this plot what two variables are anti-correlated (i.e. have minus values)?


```{r}
library(corrplot)
cij <- cor(candy)
corrplot(cij)

pairs <- which(cij == min(cij), arr.ind = TRUE)
rownames(cij)[pairs[,1]]
colnames(cij)[pairs[,2]]

```

>Q23. Similarly, what two variables are most positively correlated?


```{r}
pairs_pos <- which(cij == max(cij[upper.tri(cij)]), arr.ind = TRUE)
rownames(cij)[pairs_pos[,1]]
colnames(cij)[pairs_pos[,2]]

```


>Q24. What original variables are picked up strongly by PC1 in the positive direction? Do these make sense to you?

```{r}
pca <- prcomp(candy, scale=TRUE)
loadings_pc1 <- sort(pca$rotation[,1], decreasing = TRUE)
loadings_pc1

```

